---
title: "DATA 622 - Homework3 - Loan Approval Prediction"
author: "Diego Correa, Amanda Schettini & Soumya Ghosh"
date: "October 10, 2021"
always_allow_html: yes
output:
  html_document:
    df_print: kable
    theme: cerulean
    highlight: pygments
    css: ./lab.css
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 5
  pdf_document: default
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries

```{r warning=FALSE, message=FALSE}
library(kableExtra)
library(tidyverse)
library(ggplot2)
library(dplyr)
library(psych)
library(caret)
library(mice)
library(randomForest)

```

## Background

For this assignment, we will be working with a dataset on loan approval status. The **'Loan_Status'** is the target variable here -

### Data Dictionary
![**Loan Approval Status Data Dictionary**](C:\Users\soumya.ghosh\OneDrive - Slalom\Work Backup\CUNY\Semester7(Fall)\Homework\Homework3\Images\DataDictionary.PNG)

### Problem Statement

1. As we begin working with the dataset, we will conduct a thorough exploratory data analysis. This step is necessary as we figure out which variables should be included in models. (10 points)
2. We will use the LDA algorithm to predict the loan approval status. This will include the walk through for the steps we took, and how we decided on the key variables. (40 points)
3. Use K-nearest neighbor (KNN) algorithm to predict the species variable. Please be sure to walk through the steps you took. This includes talking about what value for ‘k’ you settled on and why. (40 points)
4. Use Decision Trees to predict on loan approval status. (40 points)
5. Use Random Forests to predict on loan approval status. (40 points)
6. Model performance: Comparison of the models we settled on in problem # 2- 5. Comment on their relative performance. Which one would you prefer the most? Why? (5 points)

## Dataset

```{r warning=FALSE, message=FALSE}
dataset <- read_csv('Loan_approval.csv')
head(dataset)%>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% 
  scroll_box(width="100%",height="300px")
```

### Descriptive Dataset Summary

```{r warning=FALSE, message=FALSE}
summary(dataset)%>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="400px")
```

## Exploratory Data Analysis

```{r message=FALSE, warning=FALSE, fig.width=10}
cat_vars <- dataset %>%
  select(-c('Loan_ID', 'ApplicantIncome', 'CoapplicantIncome','LoanAmount','Loan_Amount_Term')) %>%
  gather(key = 'predictor_variable', value = 'value', -Loan_Status)

# Plot and print a histogram for each predictor variable.
ggplot(cat_vars) +
  geom_histogram(aes(x = value, fill = Loan_Status),stat='count', bins = 30) +
  labs(title = 'Distributions of Categorical Variables') +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(. ~predictor_variable, scales = 'free', ncol = 3)
```

```{r message=FALSE, warning=FALSE, fig.width=10}
cont_vars <- dataset %>%
  select(ApplicantIncome, CoapplicantIncome, LoanAmount, Loan_Amount_Term, Loan_Status) %>%
  gather(key = 'predictor_variable', value = 'value', -Loan_Status)

# Plot and print a histogram for each predictor variable.
ggplot(cont_vars) +
  geom_histogram(aes(x = value, y = ..density.., fill = Loan_Status), bins = 30) +
  labs(title = 'Log Distributions of Continuous Variables') +
  scale_x_log10() +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(. ~predictor_variable, scales = 'free', ncol = 3)
```

## Pre-Processing

### Missing Value Analysis

```{r message=FALSE, warning=FALSE}
## Counts of missing data per feature
dataset_missing_counts <- data.frame(apply(dataset, 2, function(x) length(which(is.na(x)))))
dataset_missing_pct <- data.frame(apply(dataset, 2,function(x) {sum(is.na(x)) / length(x) * 100}))

dataset_missing_counts <- cbind(Feature = rownames(dataset_missing_counts), dataset_missing_counts, dataset_missing_pct)
colnames(dataset_missing_counts) <- c('Feature','NA_Count','NA_Percentage')
rownames(dataset_missing_counts) <- NULL


dataset_missing_counts %>% filter(`NA_Count` != 0) %>% arrange(desc(`NA_Count`)) %>% kable() %>% kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>% scroll_box(width="100%",height="300px")

ggplot(dataset_missing_counts, aes(x = NA_Count, y = reorder(Feature, NA_Count))) + 
  geom_bar(stat = 'identity', fill = 'steelblue') +
  labs(title = 'Missing Counts') +
  theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank())

```

```{r eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
dataset_missing_counts <- sapply(dataset, function(x) sum(is.na(x))) %>% 
  sort(decreasing = TRUE) %>%
  as.data.frame() %>% rename('NA_Count' ='.') 

dataset_missing_counts <- dataset_missing_counts%>%
  mutate('Feature' = rownames(dataset_missing_counts))

ggplot(dataset_missing_counts, aes(x = NA_Count, y = reorder(Feature, NA_Count))) + 
  geom_bar(stat = 'identity', fill = 'steelblue') +
  labs(title = 'Missing Counts') +
  theme(plot.title = element_text(hjust = 0.5), axis.title.y = element_blank(), axis.title.x = element_blank())
```

### Data Imputation 

```{r message=FALSE, warning=FALSE}
#transformation
#Loan_ID should be removed before imputing data
#mice uses all data to impute
#Transforming continuous variables by taking the log
dataset <- dataset %>%
  select(-'Loan_ID') %>%
  mutate(
    Gender = as.factor(Gender),
    Married = as.factor(Married),
    Dependents = as.factor(Dependents),
    Self_Employed = as.factor(Self_Employed),
    Credit_History = as.factor(Credit_History),
    ApplicantIncome = log(ApplicantIncome),
    CoapplicantIncome = log(CoapplicantIncome),
    LoanAmount = log(LoanAmount),
    Loan_Amount_Term = log(Loan_Amount_Term)
  )
```



```{r message=FALSE, warning=FALSE}
#imputation by using the random forest method ('rf')
init <- mice(dataset, maxit = 0)
predM <- init$predictorMatrix
set.seed(123)
imputed <- mice(dataset, method = 'rf', predictorMatrix = predM, m=5)
```

```{r message=FALSE, warning=FALSE}
dataset <- complete(imputed)
summary(dataset)
```

```{r message=FALSE, warning=FALSE} 
# none of the variables meet the condition to be a degenerate feature
nearZeroVar(dataset)
```


